---
title: "Handling Missing Data (Long Gaps)"
author: "Jayita Chakraborty (PhD)"
date: ""
output: 
  pdf_document:
    latex_engine: xelatex
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
header-includes:
  - \usepackage{xeCJK}
  - \usepackage{booktabs}
colorlinks: true
urlcolor: blue
---

```{r setup, include=FALSE}
library(formatR)
options(width = 60)
knitr::opts_chunk$set(echo = TRUE,tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```


## Joining Information to data
Let's use **London Smart Meters** dataset from Kaggle for the current exercise. The source of the data is from London Data Store and Jean-Michel D. has augumented the dataset with a few more details and packaged it as a Kaggle Dataset.

**About the Dataset**
London Data Store, free and open data-sharing portal, provided this dataset, which was collected and enriched by Jean-Michel D and uploaded on [Kaggle](https://www.kaggle.com/jeanmidev/smart-meters-in-london). The dataset has energy consumption readings for a sample of 5,567 London Households that took part in the UK Power Networks led Low Carbon London project between November 2011 and February 2014. Readings were taken at half hourly intervals. Some metadata about the households are also available as part of the dataset.

```{r message=FALSE, warning=FALSE}
require(readr)
require(dplyr)
require(tidyverse)
require(fs)
require(tsibble)
require(imputeTS)
require(arrow)
```



```{r}
source_data <- path("../archive")
block_data_path <- path(source_data, "hhblock_dataset", "hhblock_dataset")
```

```{r}
# Read the CSV without parsing dates automatically (like parse_dates=False)
block_1 <- read_csv(path(block_data_path, "block_0.csv"),col_types = cols(day = col_character()))
str(block_1)
```

```{r}
block_1$day <- date(block_1$day)
block_1_tibble <-  as_tsibble(block_1,key=LCLid,index = day)
```

```{r}
block_1_tibble %>%
  group_by_key() %>%
  summarise(day = max(day)) %>%
  as_tibble() %>%
  slice_sample(n=5)
```

```{r}
# Get all block file paths
block_files <- dir_ls(block_data_path, glob = "*.csv")

# Read and bind all 111 files together
all_blocks <- map_dfr(
  block_files,
  ~ read_csv(.x, col_types = cols(day = col_character()))
)

# Parse the date column once, across the combined data
all_blocks$day <- date(all_blocks$day)

# Convert the full combined dataset to a single tsibble
all_blocks_tsbl <- as_tsibble(all_blocks, key = LCLid, index = day)

# Maximum date across ALL blocks/households
print(paste("Max Date across all blocks: ", max(all_blocks_tsbl$day)))
print(paste("Max Date across all blocks: ", min(all_blocks_tsbl$day)))
```


```{r}
head(all_blocks_tsbl)
```


```{r}
all_blocks_tsbl_long <- all_blocks_tsbl %>%
  pivot_longer(
    cols = starts_with("hh_"),
    names_to = "hour_block",
    values_to = "energy_consumption"
  )

all_blocks_tsbl_long <- all_blocks_tsbl_long %>%
  mutate(
    hour_num = as.integer(gsub("hh_","",hour_block))
  )

head(all_blocks_tsbl_long)
```


```{r}
all_blocks_long_tsbl <- all_blocks_tsbl_long %>%
  mutate(timestamp = as_datetime(day) + minutes(30 * hour_num)) %>%
  as_tsibble(key = LCLid, index = timestamp)

# global max across the whole block (or dataset)
max_date <- max(all_blocks_long_tsbl$timestamp)

all_blocks_long_tsbl <- all_blocks_long_tsbl %>%
  fill_gaps(.end = max_date)

all_blocks_compact <-all_blocks_long_tsbl %>%
  as_tibble() %>%
  group_by(LCLid) %>%
  summarise(
    start_timestamp = min(timestamp),
    energy_consumption = list(energy_consumption),
    series_length = n(),
    .groups = "drop"
  ) %>%
  mutate(frequency = "30min") %>%
  select(LCLid, start_timestamp, frequency, energy_consumption, series_length)

head(all_blocks_compact)
```


```{r}
# Removing all other unused variables from memory
rm(list = c("all_blocks","all_blocks_long_tsbl","all_blocks_tsbl","all_blocks_tsbl_long","block_1","block_1_tibble"))
gc()


# Per-column memory usage (in bytes)
col_sizes <- data.frame(map_dbl(all_blocks_compact, object.size))
names(col_sizes) <- "Object_size"

print(col_sizes)

# Total size in MB
cat("Total:", sum(col_sizes) / 1024^2, "MB\n")
```
## Merging additional information
### Household information
```{r}
household_info <- read_csv(path(source_data,"informations_households.csv"))
head(household_info)
```


```{r}
hourlyblock_df <- all_blocks_compact %>%
  left_join(household_info,by = "LCLid")

head(hourlyblock_df)
```


```{r}
# remove all_blocks_compact
rm(all_blocks_compact)
gc()
```

### Add Bank Holidays
```{r}
bank_holidays <- read_csv(path(source_data,"uk_bank_holidays.csv"), col_types = cols('Bank holidays'=col_character()))
bank_holidays$`Bank holidays`<- as.Date(bank_holidays$`Bank holidays`)

head(bank_holidays)
```

```{r}
bank_holidays_full <- bank_holidays %>%
  rename(date = `Bank holidays`) %>%
  as_tsibble(index = date) %>%
  fill_gaps(.full = TRUE) %>%
  as_tibble() %>%
  mutate(Type = replace_na(Type, "NO_HOLIDAY")) %>%
  reframe(time = as_datetime(date) + minutes(30 * (0:47)), Type=Type, .by = date) %>%
  select(time, Type) %>%
  as_tsibble(index=time)

head(bank_holidays_full)
```
### Add Weather

```{r}
weather_hourly <- read_csv(path(source_data,"weather_hourly_darksky.csv"), col_types = cols(time=col_character()))
weather_hourly$time <- as_datetime(weather_hourly$time)
```

```{r}
full_time <- tibble(time = seq(min(weather_hourly$time), max(weather_hourly$time), by = "30 min"))

weather_hourly_full <- full_time %>%
  left_join(weather_hourly %>%  as_tsibble(index=time),by="time") %>%
  na_locf()

head(weather_hourly_full)
```


```{r}
map_weather_holidays <- function(row, bank_holidays, weather_hourly) {

  # Create regular timestamp sequence
      # using regular expression to add space between the numeric part 
    # and text part of the variable
  row$frequency <- gsub("([0-9]+)([a-zA-Z]+)","\\1 \\2",row$frequency)
  date_range <- seq(
    from = row$start_timestamp,
    by = row$frequency,
    length.out = row$series_length
  )

  # Empty tsibble
  std_tbl <- tibble(time = date_range) %>%
    as_tsibble(index = time)

  # Join holidays
  holidays <- std_tbl %>%
    left_join(bank_holidays, by = "time") %>%
    mutate(Type = if_else(is.na(Type), "NO_HOLIDAY", Type))

  # Join weather
  weather <- std_tbl %>%
    left_join(weather_hourly, by = "time")

  # Assertion
  stopifnot(nrow(holidays) == row$series_length)
  stopifnot(nrow(weather) == row$series_length)

  # Store holidays as vector
  row$holidays <- list(holidays$Type)

  # Store each weather variable as a vector
  weather_cols <- setdiff(names(weather), "timestamp")
  
  for (col in weather_cols) {
    row[[col]] <- list(weather[[col]])
  }

  return(row)
}
```


```{r}
hourlyblock_df_compact <- lapply(seq_len(nrow(hourlyblock_df)), function(i) {
  print(paste0("Processing row ",i))
  map_weather_holidays(
    hourlyblock_df[i, ],
    bank_holidays_full,
    weather_hourly_full
  )
})
```


```{r}
rm(list=c("bank_holidays_full","household_info","weather_hourly","weather_hourly_full","full_time"))
gc()

hhblock_tbl <- map_dfr(hourlyblock_df_compact,as_tibble)

parquet_dir <- path(source_data, "hhblock_parquet")
parquet_file <- path(
    parquet_dir,
    paste0("hourlyblock_chunks_merged", ".parquet")
  )
write_parquet(hhblock_tbl, parquet_file)

```




```{r}
dir.create(path(source_data,"hourlyblock_chunks"), showWarnings = FALSE)

# Split into 8 approximately equal chunks
chunks <- split(
  hourlyblock_df_compact,
  cut(seq_along(hourlyblock_df_compact), breaks = 8, labels = FALSE)
)

# Save each chunk for saving them to .rds format
for (i in seq_along(chunks)) {
  saveRDS(chunks[[i]], file = path(path(source_data,"hourlyblock_chunks"), sprintf("hourlyblock_df_compact_%d.rds", i)))
}
```

